Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-06-28 Manuscript.tex
more edits

Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} sp.
shells were much lower (0.05--0.30) than those seen in other patelloid
species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in
Hausmann \emph{et al}. \citeyearpar{Hausmann2023-ih}) using the same
emission lines. While the intensity ratio depends chiefly on the chosen
emission lines to calculate the proportion of the chosen Magnesium to
the chosen Calcium peak, the low ratios still indicate lower
concentrations than seen elsewhere, or seen only in aragonitic parts of
e.g.~\emph{Patella} shells. Similar observations have been made before
\citep{Graniero2017-io}.

Line scans reflect well the changes previously indicated by
\(\delta\)\textsuperscript{18}O values and also indicate about two years
of recorded growth. That said, the 2D imaging suggests that in both
specimens a third summer can be added to the total record, which is not
captured by the line scans. The variability seen in the \emph{P.
vulgata} specimens above is not as strong, which might also be related
to the fast and consistent growth rate of the shells.

@@

\caption{\label{fig-Nac_Comp}Comparison of LIBS line-scan and sequential
\(\delta\)\textsuperscript{18}O-values of specimen ORK-LT5}

@@

The Mg/Ca ratios from the 2D elemental imaging present positive results
on SST being the main factor behind Mg/Ca concentration in three
Atlantic limpet species. Annually repeating patterns of increasing
(spring/summer) and decreasing (autum/winter) SST were also visible in
the line scans, albeit less straightforward than in the 2D-data.
Nevertheless, the resulting R\textsuperscript{2}-values for each species
were high (\emph{P. vulgata mean R\textsuperscript{2}}=0.89, \emph{N.
deaureata mean R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
R\textsuperscript{2}}=0.88). These results are similar to previously
published data on \emph{Patella} sp. shells (Table
\hyperref[Table_2]{2}). R\textsuperscript{2} values of \emph{P.
depressa} lie between 0.78 and 0.87 \citep{Garcia-Escarzaga2021-ij}, for
\emph{P. caerulea} from the Mediterranean between 0.78 and 0.96 ---
although with one outlier of 0.33 (specimen MP64A), which had increased
Mg/Ca ratios in increments with rich organic contents
\citep{Hausmann2019-fi}.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}@{}}
\caption{Table showing correlations of Mg/Ca and
\(\delta\)\textsuperscript{18}O found in other studies. * indicates that
only SST data and no other geochemical data was used. ** these samples
were used to determine a shared R\textsuperscript{2}-value of
0.79.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\emph{P. depressa} & Northern Spain & LAN541 & 0.87 &
\citep{Garcia-Escarzaga2021-ij} \\
& & LAN545 & 0.86 & \\
& & LAN554 & 0.78 & \\
& & LAN559 & 0.82 & \\
\emph{P. caerulea} & Croatia & ISTPC1 & 0.9 & \citep{Hausmann2019-fi} \\
& & ISTPC2 & 0.84 & \\
& Crete & AF1911A & 0.91 & \\
& & AF3003A & 0.92 & \\
& Israel & AKKPC2 & 0.96 & \\
& & AKKPC3 & 0.89 & \\
& & FRMPC1 & 0.84 & \\
& & FRMPC2 & 0.96 & \\
& Libya & MO31A & 0.83 & \\
& & \textbf{MP64A} & 0.33 & \\
& & MP67A & 0.96 & \\
& & MP68A & 0.81 & \\
& Malta & MA10 & 0.82 & \\
& Tunisia & TUNPC1 & 0.81 & \\
& & TUNPC2 & 0.78 & \\
& Turkey & ANTPC1 & 0.95 & \\
& & ANTPC2 & 0.93 & \\
& & KIZPC1 & 0.94 & \\
& & KIZPC2 & 0.86 & \\
\emph{P. rustica} & Gibraltar & \textbf{JL1} & 0.02 &
\citep{Ferguson2011-zl} \\
& & JL2 & 0.8** & \\
\emph{P. caerulea} & Gibraltar & JM00 & 0.69** & \\
& & JM30 & 0.83** & \\
\emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
\citep{Graniero2017-io} and this study \\
\end{longtable}

@@

\emph{P. rustica} has two studied specimens (JL1 and JL2) with
R\textsuperscript{2}-values of 0.02 and 0.8, respectively
\citep{Ferguson2011-zl} (Figure~\ref{fig-Ferg}). Together with the data
from two \emph{P. caerulea} shells (JM00 and JM30 with
R\textsuperscript{2}-values of 0.69 and 0.83), the data from JL2 was
used determine a general Mg/Ca-SST relationship with an
R\textsuperscript{2}-value of 0.79. JL1 was disregarded in that
calculation, as well as the second year of growth in JM00, which
produced anomalous data. While no additional information about those
specimens is available, it seems likely that the Mg/Ca ratios from JM00,
which were interpreted as an ontogenetic trend, were potentially
influenced by the same intra-increment heterogeneity we see in some of
our shell specimens as well. Since these shells were analysed by
homogenising the carbonate powder of the m+2 layer and the more exterior
m+3 layer, these heterogeneities could potentially increased the Mg/Ca
ratio unpredictably, leading to what might look like an ontogenetic
trend in a linear dataset. Whether these anomalies affected JL1 in the
same way, albeit stronger, is not clear but it might well be possible.

@@

\caption{\label{fig-Ferg}Correlation graphs for Ferguson et al., 2011
specimens.}

@@

Another shell whose previous results we aim to explain further is
ORK-LT5, which has previously been shown to have no substantial
correlation with \(\delta\)\textsuperscript{18}O or SST in implication
\citep{Graniero2017-io}. Different to other specimens of this study, we
were able to resample this specimen and getting Mg/Ca data on a slightly
higher resolution (20--30 µm spots in a 100 µm raster compared to 50 µm
spots spaced at 150--300 µm) and using 2 dimensions, as well as using a
continuous high-resolution line-scan at (20--30 µm spots spaced at 10
µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
previously missed summer peaks at around 2--3 mm distance to the shell
edge (grey box in Figure~\ref{fig-ORK_sub}) and revealed additional
growth at the very edge (dashed box in Figure~\ref{fig-ORK_sub})
mirroring the change towards warmer temperatures at the time of
collection (August). Including these missing parts lead to a better
alignment of both records and a resulting R\textsuperscript{2}-value of
0.88.

@@

While we were not able to resample the \emph{Nacella} sp. specimens in
Graniero \emph{et al.} \citeyearpar{Graniero2017-io}, we were able to
source other modern and archaeological specimens of the same species
from Tierra del Fuego, which can serve as an indicator. Similar to other
patelloid shells, the results of the \emph{Nacella} specimens in our
study were successful with high R\textsuperscript{2}-values (0.76--0.95)
and suggest Mg/Ca as a reliable proxy for SST (Figure~\ref{fig-Nac_Comp}
and supplementaries). Why these results are different from the studied
\emph{Nacella} sp. specimens in Graniero et al.
\citeyearpar{Graniero2017-io} is not obvious, but the occasional
alignment of the Mg/Ca and \(\delta\)\textsuperscript{18}O records of
their shells (Figure 6 in \citep{Graniero2017-io} ) indicate that there
is some relationship that --- potentially due to the low sampling
resolution or intra-increment heterogeneity --- fell apart and lead to
an overall result of an uncorrelated relationship between Mg/Ca and SST.

@@

This study presents high-resolution and 2D information on Mg/Ca
concentrations in modern and archaeological patelloid shells and their
relation to sea surface temperature. The results for \emph{Patella
vulgata}, \emph{Nacella deaureata}, and \emph{Nacella magellanica} were
consistently promising with mean R\textsuperscript{2}-values of 0.89,
0.86, and 0.88 respectively.

@@

These results are particularly promising for future analyses of
patelloid shells for the study of season of collection in archaeological
sites as well as the reconstruction of sub-annual climatic conditions in
the past. Shells of the species studied here are widely available along
the European Atlantic façade and the Argentinian coastline and make up a
large part of marine mollusc remains in archaeological sites of those
regions. Improving their analysis in terms of cost and time efficiency
through the benefits of LIBS, which can be used in this context as a
screening tool for targeted \(\delta\)\textsuperscript{18}O analysis
focusing on annual minima and maxima revealed in the Mg/Ca record,
allows researchers to increase sample numbers and produce larger and
more in-depth climatic studies than previously possible.

>>>>

Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} sp.
shells were much lower (0.05--0.30) than those seen in other patelloid
species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in
Hausmann \emph{et al}. \citeyearpar{Hausmann2023-ih}) using the same
emission lines. While the intensity ratio depends chiefly on the chosen
emission lines to calculate the proportion of the chosen Magnesium to
the chosen Calcium peak, the low ratios still indicate lower
concentrations than seen elsewhere, or seen only in aragonitic parts of
e.g.~\emph{Patella} shells. Similar observations have been made before
\citep{Graniero2017-io}. Strong intra-increment heterogeneities are
visible along the exterior of the shells (m+3 layer), where Mg/Ca
intensity ratios are higher, particularly in specimen ND-1016-3
(Figure~\ref{fig-Nac_LIBS}). The step from lower to higher along the
increment is not gradual, as is the case in NM-1016-3 (see inset of
Figure~\ref{fig-Nac_LIBS} b), but rather step-wise. Sampling across this
boundary in a linear scan would most likely lead to erroneous results.
Also sampling across gradual changes as in NM-1016-3 can influence the
overall correlation graph for a specimen.

Line scans reflect well the changes previously indicated by
\(\delta\)\textsuperscript{18}O values and also indicate about two years
of recorded growth. That said, the 2D imaging suggests that in both
specimens a third summer can be added to the total record, which is not
captured by the line scans.

@@

\caption{\label{fig-Nac_Comp}Comparison of LIBS line-scan and sequential
\(\delta\)\textsuperscript{18}O-values of \emph{Nacella} sp. specimens.}

@@

The Mg/Ca ratios from the 2D elemental imaging present positive results
on SST being the main factor behind Mg/Ca concentration in three
Atlantic limpet species. Annually repeating patterns of increasing
(spring/summer) and decreasing (autum/winter) SST were also visible in
the line scans, albeit less straightforward than in the 2D-data.
Nevertheless, the resulting R\textsuperscript{2}-values for each species
were high (\emph{P. vulgata mean R\textsuperscript{2}}=0.89, \emph{N.
deaureata mean R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
R\textsuperscript{2}}=0.88). These results are similar to previously
published data on \emph{Patella} sp. shells (Table
\hyperref[Table_2]{2}): R\textsuperscript{2} values of \emph{P.
depressa} lie between 0.78 and 0.87 \citep{Garcia-Escarzaga2021-ij}, for
\emph{P. caerulea} from the Mediterranean between 0.78 and 0.96 ---
barring one outlier of 0.33 (specimen MP64A), which had increased Mg/Ca
ratios in increments with rich organic contents \citep{Hausmann2019-fi}.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1562}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1094}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2031}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4062}}@{}}
\caption{Correlations of Mg/Ca with \(\delta\)\textsuperscript{18}O and
SST found in other studies. Bold R\textsuperscript{2} values are
interpreted as outliers. * indicates that only SST data and no other
geochemical data was used. ** these samples were used to determine a
shared R\textsuperscript{2}-value of 0.79.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\emph{P. depressa} & Northern Spain & LAN541 & 0.87 &
\citep{Garcia-Escarzaga2021-ij} \\
& & LAN545 & 0.86 & \\
& & LAN554 & 0.78 & \\
& & LAN559 & 0.82 & \\
\emph{P. caerulea} & Croatia & ISTPC1 & 0.9 & \citep{Hausmann2019-fi} \\
& & ISTPC2 & 0.84 & \\
& Crete & AF1911A & 0.91 & \\
& & AF3003A & 0.92 & \\
& Israel & AKKPC2 & 0.96 & \\
& & AKKPC3 & 0.89 & \\
& & FRMPC1 & 0.84 & \\
& & FRMPC2 & 0.96 & \\
& Libya & MO31A & 0.83 & \\
& & \textbf{MP64A} & \textbf{0.33} & \\
& & MP67A & 0.96 & \\
& & MP68A & 0.81 & \\
& Malta & MA10 & 0.82 & \\
& Tunisia & TUNPC1 & 0.81 & \\
& & TUNPC2 & 0.78 & \\
& Turkey & ANTPC1 & 0.95 & \\
& & ANTPC2 & 0.93 & \\
& & KIZPC1 & 0.94 & \\
& & KIZPC2 & 0.86 & \\
\emph{P. rustica} & Gibraltar & \textbf{JL1} & \textbf{0.02} &
\citep{Ferguson2011-zl} \\
& & JL2 & 0.8** & \\
\emph{P. caerulea} & Gibraltar & JM00 & 0.69** & \\
& & JM30 & 0.83** & \\
\emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
\citep{Graniero2017-io} and this study \\
\end{longtable}

@@

\emph{P. rustica} has two studied specimens (JL1 and JL2) with
R\textsuperscript{2}-values of 0.02 and 0.8, respectively
\citep{Ferguson2011-zl} (Figure~\ref{fig-Ferg}). Together with the data
from two \emph{P. caerulea} shells (JM00 and JM30 with
R\textsuperscript{2}-values of 0.69 and 0.83), the data from JL2 was
used determine a general Mg/Ca-SST relationship with an
R\textsuperscript{2}-value of 0.79. JL1 was disregarded in that
calculation, as well as the second year of growth in JM00, which
produced anomalous data. While no additional information about those
specimens is available, it seems likely that discarded the Mg/Ca ratios
from JM00, which were interpreted as an ontogenetic trend, were
potentially influenced by the same intra-increment heterogeneity we see
in some of our shell specimens as well. Since these shells were analysed
by homogenising the carbonate powder of the m+2 layer and the more
exterior m+3 layer, these heterogeneities could potentially increased
the Mg/Ca ratio unpredictably, leading to what might look like an
ontogenetic trend in a linear dataset. Whether these anomalies affected
JL1 in the same way, albeit stronger, is not clear but it might well be
possible.

@@

\caption{\label{fig-Ferg}Correlation graphs for Ferguson et al., (2011)
specimens.}

@@

Another shell whose results we aim to explain further is ORK-LT5, which
has previously been shown to have no substantial correlation with
\(\delta\)\textsuperscript{18}O or SST in implication
\citep{Graniero2017-io}. Different to other specimens of this study, we
were able to resample this specimen and to acquire Mg/Ca data on a
slightly higher resolution (20--30 µm spots in a 100 µm raster compared
to 50 µm spots spaced at 150--300 µm) and using 2 dimensions, as well as
using a continuous high-resolution line-scan at (20--30 µm spots spaced
at 10 µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
previously missed summer peaks at around 2--3 mm distance to the shell
edge (grey box in Figure~\ref{fig-ORK_sub}) and revealed additional
growth at the very edge (dashed box in Figure~\ref{fig-ORK_sub})
mirroring the change towards warmer temperatures at the time of
collection (August). Including these missing parts lead to a better
alignment of both records and a resulting R\textsuperscript{2}-value of
0.88 (Figure~\ref{fig-ORK_sub}).

@@

We were not able to resample the \emph{Nacella} sp. specimens in
Graniero \emph{et al.} \citeyearpar{Graniero2017-io}, which had
similarly unrelated records of \(\delta\)\textsuperscript{18}O and Mg/Ca
and pointed to no relationship between elemental ratios and SST at all.
However, we were able to source other modern specimens of the same
species from Tierra del Fuego, which shall serve here as an indicator.
Similar to other patelloid shells, the results of the \emph{Nacella}
specimens in our study were successful with high
R\textsuperscript{2}-values (0.76--0.95) and suggest Mg/Ca as a reliable
proxy for SST (Figure~\ref{fig-Nac_Comp} and supplementaries). Why these
results are different from the studied \emph{Nacella} sp. specimens in
Graniero \emph{et al}. \citeyearpar{Graniero2017-io} is not obvious, but
the occasional alignment of the Mg/Ca and
\(\delta\)\textsuperscript{18}O records of their shells (Figure 6 in
Graniero \emph{et al}. \citeyearpar{Graniero2017-io}) indicate that
there is some relationship that --- potentially due to the low sampling
resolution or intra-increment heterogeneity --- is not perfectly
visible, leading to an overall result of an uncorrelated relationship
between Mg/Ca and SST.

@@

This study presents high-resolution and 2D information on Mg/Ca
concentrations in modern and archaeological patelloid shells and their
relationship to sea surface temperature. The results for \emph{Patella
vulgata}, \emph{Nacella deaureata}, and \emph{Nacella magellanica} were
consistently promising with mean R\textsuperscript{2}-values of 0.89,
0.86, and 0.88 respectively.

@@

These results are particularly promising for future analyses of
patelloid shells for the study of season of collection in archaeological
sites as well as the reconstruction of sub-annual climatic conditions in
the past. Shells of the species studied here are widely available along
the European Atlantic façade and the Argentinian coastline and make up a
significant part of marine mollusc remains in archaeological sites of
those regions
\citep{Colonese2011-ab, Villagran2011-ld, Zangrando2016-yl}. Improving
their analysis in terms of cost and time efficiency through the benefits
of LIBS, which can be used in this context as a screening tool for
targeted \(\delta\)\textsuperscript{18}O analysis focusing on annual
minima and maxima revealed in the Mg/Ca record, will allow researchers
to increase sample numbers and produce larger and more in-depth climatic
studies than previously possible.
